---
title: "Translation Hoang - Kostrzewski"
output: html_notebook
---

Initial case study to develop methodology - translation and comparisong
of Hoang et al (human) and Kostrzewski et al (MPS)

# Load libraries and functions

Additional libraries are called internally

```{r}
library(decoupleR)
library(OmnipathR)
library(caret)
library(ropls)
library(matrixStats)

dir_root <- "E:/Jose Luis/Documents/GitHub/FattyLiverModeling/"



source(paste0(dir_root,"utils/plotting_functions.R"))
source(paste0(dir_root,"modeling/functions_translation_jose.R"))
source(paste0(dir_root,"utils/wrapper_DECOUPLER.R"))

options(expressions = 5e5)
set.seed(123)
dir_data <- paste0(dir_root,"data/")

```

# Load datasets

Focusing on Hoang et al and Kostrzewski et al. Pre-processing is intersecting features, log CPM normalization
centering, and PCA for visualization although PCA is not crucial

Doing a more stringent threshold to keep variable genes in all datasets. COuld relax to keep variable
genes in at least one datasets, but this allows for faster execution for testing

```{r}
dataset_names <- c("Hoang", "Kostrzewski", "Wang")
ref_dataset <- "Hoang"
target_dataset <- "Kostrzewski"
# Load
data_list <- load_datasets(dataset_names, dir_data = dir_data)
# Run PCA
tmp <- process_datasets(data_list, filter_variance = F)
data_list <- tmp$data_list
plt_list <- tmp$plt_list
# Define matrices of interest
Yh <- data_list$Hoang$metadata$nafld_activity_score #data_list$Hoang$metadata$Fibrosis_stage# 
Xh <- data_list[[ref_dataset]]$data_center %>% t()
Xm <- data_list[[target_dataset]]$data_center %>% t()
exp_factors <- exp_factors <- c("NPC", "Background", "TGF", "LPS",  "Cholesterol", "Fructose") # TO DO: Add to dataset itself for MPS
```

# Build PLSR model

Number of latent variables is selected internally based on 10-fold CV as the point where the 
Q2Y metric is maximal

```{r}
### Load idx for CV splits
#idx_train_CV <- createFolds(Yh, k = 10, returnTrain = T)
#save(idx_train_CV, file = paste0(dir_data,"idx_train_CV_NAFLD.RData"))
load(paste0(dir_data,"idx_train_CV_NAFLD.RData"))

### Train PLSR - lots of internal CV, so it is slow - try to make parallel
#plsr_mod <- train_plsr(Xh, Yh, idx_train_CV)
#save(plsr_mod, file = paste0(dir_data,"temp_PLSR_mod.RData"))
# Model coefficients (including intercept) - TO DO: get rid of intercept by centering Y
#Wh <- plsr_mod$Wf

#### Skipping for now to move quickly
plsr_mod <- opls(x = Xh, 
                 y = Yh,
                 predI = 3,
                 crossvalI = 5,
                 fig.pdfC = "none",
                 info.txtC = "none")

Wh <- matrix(data = 0, ncol = ncol(plsr_mod@weightMN), nrow = ncol(Xh))
rownames(Wh) <- colnames(Xh)
colnames(Wh) <- colnames(plsr_mod@weightMN)
for (ii in 1:nrow(plsr_mod@weightMN)){
  Wh[rownames(plsr_mod@weightMN)[ii], ] <- plsr_mod@weightMN[ii,]
}
######
# Get regression coefficients
Bh <- matrix(lm(Yh ~ Xh %*% Wh) %>% coef(), ncol = 1)

# Get variances in LVs
plsr_var_plsr <- colVars(Xh %*% Wh)
plsr_per_var_plsr <- round(100*plsr_var_plsr/sum(data_list$Hoang$pca$sdev^2),2)

### Visualize
# Plot predictions in full training
plt_PLSR_mod <- plot_scatter(x = Yh, y = cbind(1, Xh %*% Wh) %*% Bh, 
                                        xlab = "NAS Measured", ylab = "NAS Predicted", xjitter = T) +
                            geom_abline(slope = 1, intercept = 0, color = "indianred",
                                        linetype="dotted", size = size_line)




plt_PLSR_score <- plot_scatter(x = Xh %*% Wh[,1], y = Xh %*% Wh[,2],
                         xlab = paste0("LV1 (", plsr_per_var_plsr[1],"%)"),
                         ylab = paste0("LV2 (", plsr_per_var_plsr[2],"%)"),
                         args_extra = data.frame(color = Yh),
                         aes_extra = aes(color = color)) +
            scale_color_viridis_c() + labs(color = "NAS")

```

# Check information preservation

See how model prediction worsens when we use the MPS space

```{r}
# Project human data onto key components of MPS and back project
L <-  1:ncol(data_list[[target_dataset]]$pca$rotation)
Wm <- data_list[[target_dataset]]$pca$rotation[,L]
# Get plots - TO DO: Extend statistics with CV
plt_info_loss <- get_info_loss(Xh, Yh, Wh, Wm, Bh)
```

# Find translatable latent variables

Find latent variables that are combinations of MPS PCs (i.e. vectors of non-zero variance)
that are the most translatable

```{r}
Wm_new <- get_translatable_LV(Xh, Yh, Wh, Wm, Bh, find_extra = F)
Wm_new <- Wm_new$Wm_new

# Plot Yh predicted with this reduced space vs the larger MPS space
Ypred1 <- cbind(1, Xh %*% Wm %*% t(Wm) %*% Wh) %*% Bh
Ypred2 <- cbind(1, Xh %*% Wm_new %*% t(Wm_new) %*% Wh) %*% Bh

plt_new_basis <- plot_scatter(x = Ypred1, y = Ypred2, 
                              xlab = "NAS full MPS space", ylab = "NAS reduced space") +
                  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "indianred") +
                  stat_cor(aes(label = after_stat(r.label)), show.legend = F, color = "black")

# Find experimental factors that contribute to the new latent variable (to do: Wrap in functions)

aov(y ~., data = data.frame(y = as.numeric(Xm %*% Wm_new[,1]), 
                            data_list$Kostrzewski$metadata[, exp_factors])) %>% summary()

# Project MPS data on this space
plt_new_basis_project <- data.frame(y = as.numeric(Xm %*% Wm_new[,2]),
                                    data_list$Kostrzewski$metadata) %>%
                          ggplot(aes(x = y, y = reorder(Combination_name, y), fill = TGF)) +
                            geom_boxplot(size = size_col) +
                            facet_grid(rows = vars(NPC), cols = vars(Background)) +
                            labs(x = "LV1 score", y = NULL) 

plt_new_basis_project <- add_theme(plt_new_basis_project) + scale_fill_brewer(palette = color_palette)
```

# Find additional latent variables

Find latent variables that, if included, would help retrieve lost information for
translation. Importantly, these new latent variables can be constructed to have zero 
variance in the MPS dataset

Number of additional vectors could be obtained with cross validation

```{r}
Wm_opt <- get_translatable_LV(Xh, Yh, Wh, Wm, Bh, find_extra = T)
Wm_opt <- Wm_opt$Wm_new

# Extend latent variables
Wm_tot <- cbind(Wm_new, Wm_opt)

# Visualize
Ypred1 <- cbind(1, Xh %*% Wm_tot %*% t(Wm_tot) %*% Wh) %*% Bh
Ypred2 <- cbind(1, Xh %*% Wm_new %*% t(Wm_new) %*% Wh) %*% Bh


df <- rbind(data.frame(x = Yh, y = Ypred2, Basis = "MPS"),
           data.frame(x = Yh, y = Ypred1, Basis = "Expanded"))


plt_aug <- plot_scatter(x = df$x, 
             y = df$y,
             xlab = "NAS measured",
             ylab = "NAS predicted",
             aes_extra = aes(color = Basis),
             args_extra = list(Basis = df$Basis)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "indianred") +
  lims(x = c(min(Yh),max(Yh)), y = c(min(Yh),max(Yh))) +
  facet_wrap(facets = vars(Basis), nrow = 1) +
  stat_cor(aes(label = after_stat(r.label)), show.legend = F, color = "black")

```

# Visualize human data on latent MPS space

```{r}
Thm <- Xh %*% Wm_tot
plt_NAS_LV <- plot_scatter(x = Thm[,1], y = Thm[,3], xlab = "MPS LV 1", ylab = "MPS LV extra 1",
                           args_extra = data.frame(NAS = Yh), aes_extra = aes(color = NAS)) +
              scale_color_viridis_c()
```


## Interpret loadings of new latent variables - pathway and TF activities

TO DO: Use PCGSEA for further interpretation

```{r}
pwy_acts_LV <- get_pwy_activities(Wm_tot)
tf_acts_LV <- get_TF_activities(Wm_tot)


# Plot
ggplot(pwy_acts_LV$TABLE, aes(y = reorder(source, score), x = score)) + 
    geom_bar(aes(fill = score), stat = "identity", color = "black", show.legend = F) +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) +
    labs(x = "Pathway score", y = NULL) +
    facet_wrap(facets = vars(condition), nrow = 1, scales = "free")

# Plot pathway scores in PC loadings in Hoang data

# nice red #67001F
corrplot::corrplot(pwy_acts_extra$MAT %>% scale(),
                   p.mat = pwy_acts_extra$MAT_pval,
                   sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.9,
                   insig = 'label_sig', is.corr = F, method = "color", 
                   col=colorRampPalette(c("#053061","white","indianred"))(100), 
                   tl.col = "black", type = "full", col.lim = c(-3,3), cl.offset = 0.2,
                   cl.align.text = "l", cl.length = 3, cl.ratio = 0.2, cl.pos = "b")



loading_perturb_viper <- run_viper(tf_acts_extra$MAT, resp_net, minsize = 5)


```

# Analyze pathway activity at sample level

Bug with PCA plot and metadata, so manual for now
```{r}

pwy_acts_MPS <- get_pwy_activities(data_list$Kostrzewski$data_center)
pwy_acts_Human <- get_pwy_activities(data_list$Hoang$data_center)


plt_pca_pwy_Human <- plot_pca(pwy_acts_Human$MAT, scale_data = F, return_data = T)
plt_dummy <- plot_scatter(x = plt_pca_pwy_Human$data$x[,1], y = plt_pca_pwy_Human$data$x[,2],
                          xlab = "PC1 (60.1%)", ylab = "PC2 (13.5%)", aes_extra = aes(color = color),
                          args_extra = data.frame(color = Yh)) +
            labs(color = "NAS") + scale_color_viridis_c()

plt_dummy <- plot_scatter(x = as.numeric(pwy_acts_Human$MAT["JAK-STAT", ]), y = Yh)

plsr_pwy <- opls(x = pwy_acts_Human$MAT %>% t(), 
                   y = Yh,
                   scaleC = "center",
                   crossvalI = 5,
                   predI = NA)


plt_pca_pwy_MPS <- plot_pca(pwy_acts_MPS$MAT, scale_data = F, return_data = T)
plt_dummy <- plot_scatter(x = plt_pca_pwy_MPS$data$x[,1], y = plt_pca_pwy_MPS$data$x[,2],
                          xlab = "PC1 (78.9%)", ylab = "PC2 (9.12%)", aes_extra = aes(color = color),
                          args_extra = data.frame(color = data_list$Kostrzewski$metadata$TGF)) +
            labs(color = "TGFb added") + scale_color_brewer(palette = "Dark2")


```